---
title: "Cognitive influences in language evolution: Study 4"
output: 
  pdf_document:
    toc: true
---

```{r echo=F,eval=F}
try(setwd("~/Box Sync/papersonthego/LOAN_WORDS/VERYNEWANALYSES/"))
try(setwd("~/Documents/MPI/MonaghanAoA/Stats 2/analysis/"))
```


# Load libraries

```{r message=F,warning=F}
library(mgcv)
library(sjPlot)
library(lattice)
library(ggplot2)
library(dplyr)
library(party)
library(lmtest)
library(gridExtra)
library(scales)
library(itsadug)
library(ggfortify)
library(factoextra)
library(gridExtra)
library(reshape2)
library(binom)
```

Extra helper functions:

```{r}
source("GAM_derivaties.R")
logit2per = function(X){
  return(exp(X)/(1+exp(X)))
}

rescaleGam = function(px, n, xvar, xlab="",breaks=NULL,xlim=NULL){
  y = logit2per(px[[n]]$fit)
  x = px[[n]]$x *attr(xvar,"scaled:scale") + attr(xvar,"scaled:center")
  se.upper = logit2per(px[[n]]$fit+px[[n]]$se)
  se.lower = logit2per(px[[n]]$fit-px[[n]]$se)
  dx = data.frame(x=x,y=y,ci.upper=se.upper,ci.lower=se.lower)
  plen = ggplot(dx, aes(x=x,y=y))+
    geom_ribbon(aes(ymin=ci.lower,ymax=ci.upper), alpha=0.3)+
    geom_line(size=1) +
    xlab(xlab)+
    ylab("Probability of borrowing")
  if(!is.null(breaks)){
    plen = plen + scale_x_continuous(breaks = breaks)
  }
  if(!is.null(xlim)){
   plen = plen + coord_cartesian(ylim = c(0,1),xlim=xlim)
  } else{
    plen = plen + coord_cartesian(ylim = c(0,1))
  }
  return(plen)
}
```

# Load data

```{r}
dataloan <- read.csv("../data/loanword12.csv",stringsAsFactors = F)
dataloan$bor15 <- ifelse(dataloan$borrowing==1,1, ifelse(dataloan$borrowing==5,0,NA))
dataloan$bor15.cat <- factor(dataloan$bor15)

dataloan$subtlexzipf = as.numeric(dataloan$subtlexzipf)
dataloan$AoA = as.numeric(dataloan$AoA)
dataloan$conc = as.numeric(dataloan$conc)
dataloan$old.english.length = as.numeric(dataloan$old.english.length)
aoaSD = sd(dataloan$AoA,na.rm = T)
aoaMean = mean(dataloan$AoA/aoaSD,na.rm=T)
dataloan$cat = factor(dataloan$cat)
dataloan$effect = factor(dataloan$effect)
```

Select only complete cases.

```{r}
dataloan2 = dataloan[complete.cases(dataloan[,
               c("phonlength","AoA",
               "subtlexzipf", "cat",
               'conc','bor15')]),]
```

Scale and center:

```{r}
dataloan2$AoAscale <- scale(dataloan2$AoA)
dataloan2$subtlexzipfscale <- scale(dataloan2$subtlexzipf)
phonlength.center = median(dataloan2$phonlength)
dataloan2$phonlengthscale <-
  dataloan2$phonlength - phonlength.center
phonlength.scale = sd(dataloan2$phonlengthscale)
dataloan2$phonlengthscale = dataloan2$phonlengthscale/phonlength.scale
attr(dataloan2$phonlengthscale,"scaled:scale") = phonlength.scale
attr(dataloan2$phonlengthscale,"scaled:center") = phonlength.center
dataloan2$concscale <- scale(dataloan2$conc)
dataloan2$cat = relevel(dataloan2$cat,"Noun")
dataloan2$AoA_objscaled = scale(dataloan2$AoA_obj)
```


# GAM for Old English

With old.english length as length. The idea of this analysis is to investigate whether length prior to borrowing affected whether the word was borrowed, rather than length of the borrowing. For that we need to go back to how the language was before the borrowings

Select only complete cases with old english length also:
```{r}
dataloan3 = dataloan2[complete.cases(dataloan2[,
                                               c("phonlength","AoA",
                                                 "subtlexzipf", "cat",
                                                 'conc','bor15','old.english.length')]),]
```

Scale length:
```{r}
old.english.length.center = median(dataloan3$old.english.length)
dataloan3$old.english.lengthscale <-
  dataloan3$old.english.length - old.english.length.center
old.english.length.scale = sd(dataloan3$old.english.lengthscale)
dataloan3$old.english.lengthscale = dataloan3$old.english.lengthscale/old.english.length.scale
attr(dataloan3$old.english.lengthscale,"scaled:scale") = 
  old.english.length.scale
attr(dataloan3$old.english.lengthscale,"scaled:center") = old.english.length.center

```

Run GAM:

```{r}
m0oe = bam(bor15.cat ~
           s(old.english.lengthscale) + 
           s(AoAscale) + 
           s(subtlexzipfscale) +
           s(concscale) +
           s(cat,bs='re')+
           s(cat,old.english.lengthscale,bs='re')+
           s(cat,AoAscale,bs='re')+
           s(cat,subtlexzipfscale,bs='re')+
           s(cat,concscale,bs='re'),
         data = dataloan3,
         family='binomial')
summary(m0oe)
```

There is a significant interaction between part of speech (`cat`) and old english length, but no significant effect of old english length on its own.

Test effect of old english length without interactions:

```{r}
m1oe = bam(bor15.cat ~
             s(old.english.lengthscale) + 
             s(AoAscale) + 
             s(subtlexzipfscale) +
             s(concscale) +
             s(cat,bs='re'),
           data = dataloan3,
           family='binomial')
summary(m1oe)
```

Now Old English length is significant and linear, as for Modern English phonlength. But we are still including words that may have been borrowed BEFORE Old English, so select only those borrowed after Old English:

## Words borrowed after Old English

Classify as borrowed only those words borrowed after 900CE (so have changed since old english)


```{r}
dataloan3$age_oldest_num <- as.numeric(dataloan3$age_oldest_num)
dataloan3$bor15oe <- ifelse(dataloan3$bor15==1 & 
                              dataloan3$age_oldest_num <=1100 ,
                            1, ifelse(dataloan3$bor15==0,0,NA))
dataloan3$bor15oe.cat <- factor(dataloan3$bor15oe)
```

Run GAM:

```{r}
m2oe = bam(bor15oe.cat ~
             s(old.english.lengthscale) + 
             s(AoAscale) + 
             s(subtlexzipfscale) +
             s(concscale) +
             s(cat,bs='re')+
             s(cat,old.english.lengthscale,bs='re')+
             s(cat,AoAscale,bs='re')+
             s(cat,subtlexzipfscale,bs='re')+
             s(cat,concscale,bs='re'),
           data = dataloan3,
           family='binomial')
summary(m2oe)
```

Again, interaction between cat and length, so remove interactions to investigate main effects:

```{r}
m3oe = bam(bor15oe.cat ~
             s(old.english.lengthscale) + 
             s(AoAscale) + 
             s(subtlexzipfscale) +
             s(concscale) +
             s(cat,bs='re'),
           data = dataloan3,
           family='binomial')
summary(m3oe)
```

As before, a significant linear effect of length.

This is evidence that length of the Old English form influenced the likelihood of that word subsequently being borrowed, and is not just a consequence of borrowed words coming from languages with longer word length than English.

## Effects for each part of speech

What is the relationship between old english word length and probability of borrowing for each part of speech?  Below are plots which are calculated by using the model `m2oe` to predict borrowing probaility for a range of combinations of all variables, but keeping `cat` (part of speech) fixed to e.g. Nouns. Then the average borrowing probaility for each value of old english word length is plotted.  The code is hidden, but can be seen in the Rmd file.  These are for diagnosis only, formal tests are below.

```{r echo=F}
plotLengthEffectForPoS = function(posType="Noun",add=F,lineCol=1){
  dx = dataloan3
  # Create new data evenly across range
  predictData = expand.grid(
    concscale = seq(min(dx$concscale),
                    max(dx$concscale),
                    length.out = 10),
    AoAscale = seq(min(dx$AoAscale),
                    max(dx$AoAscale),
                    length.out = 10),
    subtlexzipfscale = seq(min(dx$subtlexzipfscale),
                    max(dx$subtlexzipfscale),
                    length.out = 10),
    old.english.lengthscale = sort(unique(dx$old.english.lengthscale)),
    cat = c(posType)
  )
  # Predict probaility of each category
  pred = predict(m2oe,newdata = predictData)
  
  borVals = tapply(pred,predictData$old.english.lengthscale,mean)
  # Confidence intervals
#   borVals.sem = tapply(pred,
#                   predictData$old.english.lengthscale,
#                   function(X){sd(X)/sqrt(length(X))})
#   borVals.ciL = borVals-(2*borVals.sem)
#   borVals.ciU = borVals+(2*borVals.sem)
  # Rescale
  borVals = logit2per(borVals)
  #borVals.ciL = logit2per(borVals.ciL)
  #borVals.ciU = logit2per(borVals.ciU)
  
  lenVals = tapply(predictData$old.english.lengthscale,
                  predictData$old.english.lengthscale,mean)
  # Rescale
  lenVals = lenVals * old.english.length.scale + old.english.length.center
  if(add){
   points(lenVals,borVals,type='l',col=lineCol) 
  } else{
  plot(lenVals,borVals,type='l',
       ylim=c(0,1),
       xlab="Old English word length",
       ylab="Probability of borrowing",
       main="")
  }
  #points(lenVals,borVals.ciU,type='l',lty=2)
  #points(lenVals,borVals.ciL,type='l',lty=2)
}
plotLengthEffectForPoS("Noun")
plotLengthEffectForPoS("Verb",T,2)
plotLengthEffectForPoS("Adjective",T,3)
legend(2,1,legend=c("Nouns","Verbs","Adjectives"),lty=1,col=1:3,ncol = 3)
```


Analyse nouns only: Old English length is significant and monotonically increasing for nouns.

```{r}
m2oen = bam(bor15oe.cat ~
              s(old.english.lengthscale) + 
              s(AoAscale) + 
              s(subtlexzipfscale) +
              s(concscale),
            data = dataloan3,subset = cat == "Noun",
            family='binomial')
summary(m2oen)
plotGAMSignificantSlopes(m2oen,"old.english.lengthscale","OE Length")
```

Analyse adjectives: Old English length is significant and increasing for adjectives.

```{r}
m2oea = bam(bor15oe.cat ~
              s(old.english.lengthscale, k = 7) + 
              s(AoAscale) + 
              s(subtlexzipfscale) +
              s(concscale),
            data = dataloan3,subset = cat == "Adjective",
            family='binomial')
summary(m2oea)
plotGAMSignificantSlopes(m2oea,"old.english.lengthscale","OE Length")
```

Analyse verbs: Length is not significantly related to the probability of borrowing for verbs.

```{r}
m2oev = bam(bor15oe.cat ~
              s(old.english.lengthscale, k = 7) + 
              s(AoAscale) + 
              s(subtlexzipfscale) +
              s(concscale),
            data = dataloan3,subset = cat == "Verb",
            family='binomial')
summary(m2oev)
```



# Model plots

Plot the model estimates with Old English length, changing the dependent scale to probability and the independent variables to their original scales. This code is hidden, but you can view it in the Rmd file.

Plotting model estimates:

```{r echo=F, fig.show="hide"}
px = plot.gam(m3oe,select=1, xlab="Old English Word length", ylab="Log odds of borrowing",shade = T)
#px = plot.gam(m3oe,select=2, xlab="AoA", ylab="Log odds of borrowing",shade = T)
#px = plot.gam(m3oe,select=3, xlab="Frequency", ylab="Log odds of borrowing",shade = T)
```

```{r echo=F}
y = logit2per(px[[1]]$fit)
x = px[[1]]$x *old.english.length.scale + old.english.length.center
se.upper = logit2per(px[[1]]$fit+px[[1]]$se)
se.lower = logit2per(px[[1]]$fit-px[[1]]$se)
dx = data.frame(x=x,y=y,ci.upper=se.upper,ci.lower=se.lower)
poelen = ggplot(dx, aes(x=x,y=y))+
  geom_ribbon(aes(ymin=ci.lower,ymax=ci.upper), alpha=0.3)+
  geom_line(size=1) +
  xlab("Old English Word Length")+
  ylab("Probability of borrowing")+
  coord_cartesian(ylim = c(0,1))
paoa = rescaleGam(px,2,dataloan2$AoAscale, "Age of acquisition")
pfreq = rescaleGam(px,3,dataloan2$subtlexzipfscale, "Frequency")
grid.arrange(poelen,pfreq,paoa, nrow=1)

```

\newpage

# Swadesh words

Next analyses are for likelihood of borrowing just for the Swadesh words:

Identify Swadesh words:

```{r}
swd = read.csv("../data/SwadeshConcepts.txt", 
               header = F, stringsAsFactors = F)$V1
dataloan3$Swadesh = dataloan3$word %in% swd
```

Run GAM:

The phonetic length variable only has 10 unique values, leading to fewer unique covariate combinations than the default degrees of freedom. Following Winter & Wieling (2016), we set the number of knots to half of the unique values (k=5).

```{r}
m0sw = bam(bor15.cat ~
             s(phonlengthscale,k=5) + 
             s(AoAscale) + 
             s(subtlexzipfscale) +
             s(concscale) +
             s(cat,bs='re'),
           data = dataloan3,
           subset = Swadesh == TRUE,
           family='binomial')
summary(m0sw)
```

The only significant effect is length.

```{r}
px = plot.gam(m0sw,select=1, xlab="Length (Swadesh words)", ylab="Log odds of borrowing",shade = T)
```

\newpage

# Coexistence, insertion or replacement

This works out influences on type of borrowing: coexistence, insertion, or replacement:

```{r}
dataloan2$effect2 = dataloan2$effect
dataloan2$effect2[!dataloan2$effect2 %in%
      c("Insertion","Replacement","Coexistence")] = NA
dataloan2$effect2 = factor(dataloan2$effect2)
dataloan2$effect2 = as.numeric(dataloan2$effect2)-1
```

Multinomial GAM (predicting probability of each category given length, AoA, frequency, concreteness and PoS).

```{r}
mCIR = gam(
        list(effect2 ~
            # formula for 2nd category
             s(phonlengthscale)+ 
             s(AoAscale) + 
             s(subtlexzipfscale) +
             s(concscale) +
             s(cat,bs='re'),
            # formula for 3rd category
             ~ s(phonlengthscale)+ 
             s(AoAscale) + 
             s(subtlexzipfscale) +
             s(concscale) +
             s(cat,bs='re')),
           data = dataloan2[!is.na(dataloan2$effect2),],
           family=multinom(K=2))
summary(mCIR)
```

Only concreteness is a significant predictor.

```{r echo=F,eval=F}
par(mfrow=c(2,2))
for(i in c(1:4,6:9)){
  plot.gam(mCIR,select=i)
}
par(mfrow=c(1,1))
plot.gam(mCIR,select=4)
plot.gam(mCIR,select=9)
```

Visualise the relationship between concreteness and the probability of each category. Code is not shown, but available in the Rmd file.

```{r echo=F}
dx = dataloan2[!is.na(dataloan2$effect2),]
# Create new data evenly across range
predictData = expand.grid(
  concscale = seq(min(dx$concscale),
                  max(dx$concscale),
                  length.out = 10),
  AoAscale = seq(min(dx$AoAscale),
                  max(dx$AoAscale),
                  length.out = 10),
  subtlexzipfscale = seq(min(dx$subtlexzipfscale),
                  max(dx$subtlexzipfscale),
                  length.out = 10),
  phonlengthscale = sort(unique(dx$phonlengthscale)),
  cat = unique(dx$cat)
)
# Predict probaility of each category
pred = predict(mCIR,newdata = predictData, type="response")

# Summarise the responses for concreteness
#  across other variables
probRes = data.frame(
  concscale = 
    tapply(predictData$concscale,predictData$concscale,mean),
  probCoexistence =
    tapply(pred[,1],predictData$concscale,mean),
  probInsertion = 
    tapply(pred[,2],predictData$concscale,mean),
  probReplacement = 
    tapply(pred[,3],predictData$concscale,mean))

 #Rescale
probRes$concscale = probRes$concscale *
  attr(dataloan2$concscale,"scaled:scale") +
  attr(dataloan2$concscale,"scaled:center")

# Plot
plot(probRes$concscale,
     probRes$probCoexistence,
       type='l',lwd=2,
     ylim=c(0,1),ylab="Probability",
     xlab="Concreteness")
points(probRes$concscale,
       probRes$probInsertion,
       type='l',col=2,lwd=2)
points(probRes$concscale,
       probRes$probReplacement,
       type='l',col=3,lwd=2)
legend(2,1,
       legend = c("Coexistence","Insertion","Replacement"),
       col=1:3,lwd=2,lty=1)

```

